---
title: "Analysis of BMI and Daily Minutes of Physical Activity considering Smoking Status"
author: "Nicole Ha"
format: pdf
number-sections: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#read cvs file of cchs dataset
library(tidyverse)
cchs_dataset<-read.csv(file="C:/Users/n/Downloads/cchs (3).csv")

```

# Introduction

With a little more than a quarter of Canadian adults affected by obesity in 2018(McDiarmid, 2019), healthy lifestyle habits has become more important over the years for managing Canadians' health. Obesity is defined as a body mass index(BMI) of 30 or higher and BMI is a measure of a person's estimated body fat, with units of weight in kg divided by height in metres squared(Centers for Disease Control and Prevention,2022). BMI can also serve as a health indicator on possible disease risk(Centers for Disease Control and Prevention,2022). Lifestyle choices such as daily amount of time spent physically active and frequency of smoking contribute to a person's overall health. Physical activity, in particular, can influence the weight of an individual and accordingly, their BMI. There is also evidence that smoking negatively affects health and while studies have had mixed results, one study by Dare et al.(2015) suggested that current smoking behaviour was associated with lowers BMIs and that former smokers were more likely than lifetime abstainers and current smokers to have higher BMIs reaching the obesity range. Therefore,it can be useful to understand how BMI relates to physical activity, keeping other factors like smoking status constant. In this paper, I pose the question, What is the relationship between average daily number of minutes spent physically active and body mass index after taking into account smoking status?   

I hypothesized that a negative correlation exists between body mass index (BMI) and daily number of minutes of physical activity. My rationale is that the greater number of minutes exercising, on average, would help a person maintain a healthy weight and potentially a lower BMI. On the other hand, the null hypothesis would state that there is no correlation.
# Data

The CCHS(2017-2018),a cross-sectional survey that sampled Canadians,was used for this analysis and includes data on health determinants(Government of Canada, 2018). The CCHS picks out clusters of residences and then uses systematic sampling in which they select residences by a set interval (Government of Canada, 2018).Through stratified sampling, it uses random samples and  thus, is representative of the general Canadian population(Government of Canada, 2018).

 80028 respondents was included in this analysis. All respondents who did report values for the variables of interest were included with no duplicated data. From the data set, the variables, average daily number of minutes physically active and body mass index were used with smoking status as the covariate.
## Descriptive statistics


```{r}
#change the names of columns
colnames(cchs_dataset)<-c('province_of_residence', 'sex','marital_status','age', 'main_activity_last week','highest_education_level', 'perceived_health','satisfaction_with_life_in_general','height(meters)','weight(kgs)','body_mass_index', 'smoking_status','number_of_weekdays physically_active', 'number_of_weekend_days_physically_active', 'total_household_income','average_daily_number_of_minutes_physically_active','working_status_last week', 'immigrant_status')

#drop NAs from dataset
no_Na_data<-cchs_dataset|>
  drop_na()

#checking for number of unique values, no duplicates were found
no_Na_data|>
  unique()|>
  tally()  

```

```{r,include=FALSE} 

#find the means of average_daily_number_of_minutes_physically_active and body_mass_index
no_Na_data|>
   summarise(mean_body_mass_index=mean(body_mass_index))

no_Na_data|>
          summarise(mean_average_daily_number_of_minutes_physically_active=mean(average_daily_number_of_minutes_physically_active))

#find the medians of average_daily_number_of_minutes_physically_active and body_mass_index
no_Na_data|>
   summarise(median_body_mass_index=median(body_mass_index))
no_Na_data|>
  summarise(median_average_daily_number_of_minutes_physically_active=median(average_daily_number_of_minutes_physically_active))

#find the modes of average_daily_number_of_minutes_physically_active and body_mass_index
no_Na_data|>
   summarise(max_body_mass_index=max(body_mass_index))
   no_Na_data|>            summarise(max_average_daily_number_of_minutes_physically_active=max(average_daily_number_of_minutes_physically_active))

#find standard deviation and range
   no_Na_data|>
   summarise(sd_body_mass_index=sd(body_mass_index),
             (diff(range(body_mass_index))))
   no_Na_data|>            summarise(sd_average_daily_number_of_minutes_physically_active=sd(average_daily_number_of_minutes_physically_active),
                (diff(range(average_daily_number_of_minutes_physically_active))))
 
```

                                     
                       
Summary Statistics |   Daily Physical Activity(mins) | Body Mass Index (kg/m^2)
------------- | -------------    |-------------
Mean | 15.56 |  26.94                  
Median | 0  |26.30
Mode |  1071.40 | 54.42
Range | 1071.40 |  40.19
Standard deviation | 31.40 |5.30

For daily physical activity, the middle value or median is zero minutes but the mean is higher than that by around 15 minutes. This may imply that there are outliers that skew the distribution to the right. In contrast, the mean and median for BMI are similar and the typical BMI value is at 26kg/m^2. The data for daily physical activity is more spread out than for body mass index as evident by daily physical activity having a greater standard deviation than BMI.

```{r}
#create a scatterplot of variables of interest
ggplot(data=no_Na_data, aes(average_daily_number_of_minutes_physically_active,body_mass_index, col=smoking_status))+
geom_point()+
  theme_bw(base_size = 6)+
scale_color_viridis_d()+
  labs(title="Body mass index vs. Average daily number of minutes physically_active", subtitle="CCHS(2017-2018)", x="Average_daily_time_physically_active (mins)",y= "BMI (kg/m^2)")
```

The scatterplot shows that the majority of respondents have 0-60 mins of daily physical activity. There is variation in respondents' BMI, especially at the lower minutes of average daily physical activity closer to 0. In regards to smoking status, there appears to be a large number of lifetime abstainers, which is more than the current daily smokers.It looks like the respondents that have a high number of minutes of daily physical activity have lower BMIs, although they also seem to be in the minority for daily minutes of physical activity with the majority of respondents having a low number of minutes of daily physical activity. So, the distribution of daily physical activity might be right skewed as well. 

```{r}
#create a histogram of BMI 
ggplot(data=no_Na_data,aes(x=body_mass_index,y = ..density..))+
  geom_histogram()+
  theme_bw(base_size = 5.5)+
  
  labs(title="Body mass index", subtitle="CCHS(2017-2018)", x="body_mass_index(kg/m^2)")
```
From the histogram, the distribution of BMI looks a bit right-skewed, which means that the majority of respondents have a BMI at the lower end at around 25kg/m^2 and the probability of having a BMI around that value is high. On the other hand, the probability of having a high BMI is much smaller and the right tail is longer. 

```{r}
#create a boxplot of BMI by smoking status
ggplot(data=no_Na_data, aes(x=body_mass_index, fill=smoking_status))+
  geom_boxplot()+
  theme_bw(base_size = 5.5)+
  labs(title="Distribution of BMI by smoking status",subtitle="CCHS(2017-2018)",x="BMI  (kg/m^2)")
```

The boxplot shows very similar BMI medians for current daily smokers and current occasional smokers as well as very similar BMI medians for former occasional smokers and experimental smoker. The former daily smokers have a slightly higher median BMI than the other smoking status categories. Former occasional smokers have a higher BMI minimum value than the others.The data for each smoking status category does not look drastically skewed as the median lines tend to be close to the center of the boxes and the interquartile ranges look similar. Overall, there looks to be little differences between the BMI variability across smoking status.Due to this similarity between BMI across smoking status, the main relationship may not vary as much across smoking status categories and so, this analysis focused on the main relationship between physical activity and BMI while keeping smoking status constant.

 
```{r include=FALSE}
#check residuals by estimating a linear regression 
lr2<-lm(body_mass_index~average_daily_number_of_minutes_physically_active+smoking_status, data=no_Na_data)
#extract residuals
residual_normal<-lr2[["residuals"]]
#add residuals to data frame
checking_res_normal<-no_Na_data|>
  mutate(residual4=residual_normal)

```
```{r}
#plotting residuals to get an idea of how the predicted values align with the observed values
ggplot(data=checking_res_normal,aes(x=average_daily_number_of_minutes_physically_active,y=residual4))+geom_point()+
  labs(y="residuals",title="Residuals vs. Average daily physical activity(mins)")
```
 Residuals were plotted to explore the data and see if any changes should be made before performing the regression analysis. Checking the residuals, the variance is unequal across the x values of average daily minutes of physical activity. This can indicate heteroscedasticity and violates a Gauss-Markov assumption underlying linear regression models. Also, BMI was found to be a right-skewed distribution. To remedy all of this, both daily minutes of physical activity and BMI variable were log transformed. 1 was added to each value in the daily minutes of physical activity, in order to conduct the log transforms.   
 
                        
```{r}
#find the correlation between main x and y variables
no_Na_data|>
select(average_daily_number_of_minutes_physically_active,body_mass_index) |>
summarise(correlation  = cor(average_daily_number_of_minutes_physically_active, body_mass_index))
```
As shown by the correlation coefficient, when the average minutes of daily physical activity increases, BMI decreases and vice versa.This negative relationship is very weak as it is far away from -1.
To determine the expected change in BMI and assess how confident one can be about the associations, I conducted a multiple linear regression in the following section.


# Model

A multiple linear regression was used to examine how BMI might be associated with daily physical activity and smoking status.The explanatory variables are average daily number of minutes physically active and smoking status. The response variable is body mass index. A multiple linear regression was found to be suitable for this analysis because it provides a model of the relationship between a response variable and two or more explanatory variables, which is relevant to the research question. 


The regression model used for this project is: 
$$
Y_i = \beta_0 + \beta_1X_{i1} + \beta_2X_{i2} +\beta_3X_{i3}+\beta_4X_{i4}+\beta_5X_{i5}+\beta_6X_{i6}
$$
where

- $Y_i =$ Predicted BMI
- $X_{i1} =$ Average daily number of minutes physically active 
- $X_{i2} =$ Current Daily Smoker 
- $X_{i3} =$Current Occasional smoker
- $X_{i4} =$Experimental smoker 
- $X_{i5} =$Former daily smoker
- $X_{i6} =$Former occasional smoker 

From the EDA, it was determined that the residuals showed unconstant variance, so to give more flexibility for nonlinearity to occur, log transforms of the daily physical activity and BMI variable were conducted and then a regression summary was provided.

```{r}
#add 1 to each value of average daily number of minutes of physical activity
daily_physical_activity_plot<-no_Na_data|>
  select(average_daily_number_of_minutes_physically_active,body_mass_index,smoking_status)|>
  mutate(activity_plus_one=average_daily_number_of_minutes_physically_active+1)

leveled_data2<-daily_physical_activity_plot|>
  mutate(smoking_status=factor(smoking_status))|>
   #reordered to make lifetime abstainer reference catgory
  mutate(smoking_status_re=fct_relevel(smoking_status,"Lifetime abstainer (never smoked a whole cigarette)",after=0))|>
  mutate(log_bmi2=log(body_mass_index))|>
  #took the log of daily physical activity with 1 added to the values 
  mutate(log_average=log(activity_plus_one))
#estimated the multiple linear regression
work_log2<-lm(log_bmi2~log_average+smoking_status_re, data=leveled_data2)
summary(work_log2)

```

# Results

Interpreting the regression, a one percent increase in the minutes of daily physical activity is associated with a 0.01 percent decrease in BMI, on average,holding smoking status constant. Setting the alpha level at 0.05, the p value is < 2e-16 for the log of daily physical activity. The probability of the observed values occurring under the null hypothesis is very low and as indicated by the three asterisks, the coefficient describing the association between daily physical activity and BMI is quite significant. 

There is a 0.62% lower BMI, on average, for current daily smokers compared to lifetime abstainers (never smoked), holding physical activity constant. The p value is 0.0021, so it is unlikely that the coefficient is zero. Indicated by the two asterisks, the regression coefficient is shown to be significant. 

There is a 0.58% lower BMI, on average, for current occasional smokers compared to lifetime abstainers (never smoked), holding physical activity constant. The p value is 0.064, so the chances that the coefficient is zero with no difference is more likely. Indicated by no asterisks, the regression coefficient is shown to be not significant.  

There is a 2.17% higher BMI, on average, for experimental smokers compared to lifetime abstainers (never smoked), holding physical activity constant.The p value is <2e-16, so it is very unlikely that the coefficient is zero. The regression coefficient is shown to be significant. 

There is a 5.78% higher BMI, on average, for former daily smokers compared to lifetime abstainers (never smoked), holding physical activity constant.The p value is <2e-16, so it is very unlikely that the coefficient is zero. The regression coefficient is shown to be significant. 

There is a 2.35% higher BMI, on average, for former occasional smokers compared to lifetime abstainers (never smoked), holding physical activity constant.The p value is 5.01e-09, so it is very unlikely that the coefficient is zero. The regression coefficient is shown to be significant. 


# Discussion

Overall, the results showed that there is a small decrease in BMI associated with increases in daily minutes of physical activity, taking into account smoking status. This is a weak, negative relationship between physical activity and BMI. Comparing the smoking statuses,current daily smokers and current occasional smokers were shown to have a lower BMI than the lifetime abstainers, although this difference was small. Experimental, former daily, and former occasional smokers had a higher BMI than the lifetime abstainers. 

The results were expected for the most part as the hypothesis of an inverse relationship was supported by some evidence from the correlation coefficient and regression coefficients. Respondents who smoked less or were former smokers tended to report higher BMIs, which aligns with the current literature(Dare et al.,2015). However, the smoking status categories.was shown to have close similarity in the spread of BMI, especially for the experimental and former occasional smokers as well as for the current daily and current occasional smokers. These smoking category pairs likely have similar smoking frequency, so they also have similar plausible changes in BMI. The finding that former daily smokers, on average, had higher BMI percentages than lifetime abstainers was interesting as current smokers were associated with lower BMIs than people who never smoked. and even with their smoking history,former daily smokers still had higher BMIs than people who barely smoke or have never smoked. Future research can examine further comparisons between smoking statuses and how former smoking may correlate with BMI.

Since other factors beyond smoking status may influence BMI, there can be noise in the data. It can be beneficial to investigate how BMI may differ across other explanatory variables such as age or gender. For instance, people can experience changes to their BMI as they age and men and women tend to have different levels of body fat. In addition, a multiple linear regression analysis including variables used in this project along with novel covariates such as age could be performed in the future. It is noted that physical activity was generally low with many respondents reporting zero  minutes of daily physical activity. Further research can also look at if there are differences in daily physical activity across age and various income levels. Understanding how BMI may correlate with different social and economic variables is important in improving the health of Canadians.

# References
Centers for Disease Control and Prevention. (2022, June 3). Body mass index (BMI). Centers for Disease Control and Prevention. Retrieved April 15, 2023, from https://www.cdc.gov/healthyweight/assessing/bmi/index.html#:~:text=Body%20Mass%20Index%20(BMI)%20is,or%20health%20of%20an%20individual. 

Centers for Disease Control and Prevention. (2022, June 3). Defining adult overweight &amp; obesity. Centers for Disease Control and Prevention. Retrieved April 15, 2023, from https://www.cdc.gov/obesity/basics/adult-defining.html#:~:text=Adult%20Body%20Mass%20Index&amp;text=If%20your%20BMI%20is%20less,falls%20within%20the%20obesity%20range. 

Dare, S., Mackay, D. F., & Pell, J. P. (2015). Relationship between smoking and obesity: a cross-sectional study of 499,504 middle-aged adults in the UK general population. PloS one, 10(4), e0123579. https://doi.org/10.1371/journal.pone.0123579

Government of Canada, S. C. (2018, June 25). Canadian Community Health Survey - annual component (CCHS). Surveys and statistical programs. Retrieved April 15, 2023, from https://www23.statcan.gc.ca/imdb/p2SV.pl?Function=getSurvey&amp;Id=329241 

McDiarmid, C. (2019, June 25). Health Fact Sheets Overweight and obese adults, 2018. Overweight and obese adults, 2018. Retrieved April 15, 2023, from https://www150.statcan.gc.ca/n1/pub/82-625-x/2019001/article/00005-eng.htm 


